Automated Verification and Synthesis of Embedded Systems: Advances and Challenges

### 2017-06-21 paper.tex
added example about BMC
Signed-off-by: Lucas Cordeiro <lucasccordeiro@gmail.com>

From the practical point of view, SAT-based or SMT-based BMC procedures have been successfully applied to verify a large number of hardware and software systems, including digital circuits and single- and multi-threaded programs. Those BMC techniques were able to find subtle bugs in real digital and embedded software systems, as reported in the available literature~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. Nonetheless, the main criticism with respect to BMC techniques relies on completeness, since they are able to prove system correctness only if an upper bound \textit{k} is known, {\it i.e.}, a bound that unfolds all loops and recursive functions to their maximum possible depth.

Due to that limitation, BMC tools are typically susceptible to exhaustion of time or memory limits, when checking complex circuit-implementations or programs with loops, whose bounds are too large or cannot be statically determined. Indeed, such an issue has pushed researchers to overcame the depth limitation and propose extensions capable of proving global correctness.

@@

One promising approach to achieve completeness, in BMC techniques, is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. The main challenge, however, of such an approach relies on computing and strengthening inductive invariants from programs. In particular, loop invariants, which are computed from programs under verification, must be inductive (and not just invariant), in order to check the corresponding verification conditions, {\it i.e.}, induction cannot determine the invariance of a non-inductive assertion~\cite{Bradley07}. For instance, consider the C-code fragment shown in Figure~\ref{figure:inductive-invariant}. If we consider the IEEE floating-point standard (IEEE 754)~\cite{4610935,Goldberg91whatevery}, then  the loop invariant $x>0$ holds initially and after each iteration; actually, $x$ tends to infinity after $128$ iterations. This invariant, however, is not inductive, given that $x>0$ before an initial iteration does not ensure that $x>0$ after each iteration; if we have initially $x=0.9$, then $x<0$ after the fourth iteration. As a consequence, even if \textit{k}-induction procedures successfully compute such assertions, which are indeed invariant, those must be inductive, so that verifiers can automatically prove program correctness. In this specific example, an inductive invariant would be $x>1$, given that if we can check that if $x>1$ before the initial iteration, then also $x>1$ after \textit{k} iterations.
%
\begin{figure}
\centering
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
#include <assert.h>
int main(void) {
 float x=2;
 while (1) {
   x = ((2*x) - 1);
 }
 return 0;
}
\end{lstlisting}
\end{minipage}
\caption{Motivating example for inductive invariants.}
\label{figure:inductive-invariant}
\end{figure}

@@

>>>>

Note that BMC analyzes only bounded program runs, but generates verification conditions (VCs) that reflect the exact path in which a statement is executed, the context in which a given function is called, and the bit-accurate representation of the expressions. A verification condition is a logical formula (constructed from the bounded program and desired correctness properties) whose validity implies that the program's behaviour agrees with its specification~\cite{Bradley07}. Correctness properties in programs can be specified by the user via \textit{assert} statements or automatically generated from a specification language~\cite{Thomas01}. If all of a bounded program's VCs are valid, then the program is in compliance with its specification up to the given bound.

As an example, consider a simple C program (slightly modified from~\cite{Strichman08}) with an exponential number of paths as shown in Figure~\ref{figure:verification-condition}(a); the corresponding C program in single static assignment (SSA) form~\cite{Appel97} is shown in Figure~\ref{figure:verification-condition}(b). Note that the SSA form is an intermediate representation used by compilers to facilitate optimizations and transformations of the program code. The common property in SSA form is that every variable has only one definition in the program text. This is achieved by introducing a fresh variable from the original name ({\it e.g.}, with a subscript) such that every assignment has a unique left hand side as shown in Figure~\ref{figure:verification-condition}(b).
%
\begin{figure}[ht]
\centering
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
#include <assert.h>
int x[N], a;
...
int main(void) {
  a=N;
  for(int i=0; i<N; i++)
    if (x[i]>1)
      a--;
  assert(a<=N);
  return 0;
}
\end{lstlisting}
\end{minipage}
\begin{center}
(a)
\end{center}
\centering
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
a1 = N
a2 = (x[0] > 1) ? a1 - 1 : a1
a3 = (x[1] > 1) ? a2 - 1 : a2
a4 = (x[2] > 1) ? a3 - 1 : a3
...
an+1 = (x[N-1] > 1) ? an - 1 : an 
\end{lstlisting}
\end{minipage}
\begin{center}
(b)
\end{center}
\caption{(a) A simple C program with a \textit{for} loop. (b) The corresponding unwound C program of (a) converted into SSA form.} 
\label{figure:verification-condition}
\end{figure}

Apart from that, this program has an exponential number of paths since each element of array $x$ can be either greater than one or less than or equal to one. Despite the large number of paths through the program, BMC unwinds it up to a bound $k$ and translates it into a VC $\psi$ such that $\psi$ is satisfiable if and only if the assertion $\left(a<=N\right)$ fails. Note that BMC still encodes the states of the program with a size that grows linearly with N. More precisely, the program in Figure~\ref{figure:verification-condition}(a) is converted into the $\psi$ using a decidable fragment of first-order logic as follows~\cite{Bradley07}:
%
\begin{equation}              
\label{verification-condition-example}
\psi := \left [ \begin{array}{ll} 
                a_{1} = N  \\
                \wedge \, \, a_{2} = ite\left(x[0] > 1, a_{1} - 1, a_{1}\right) \\ 
                \wedge \, \, a_{3} = ite\left(x[1] > 1, a_{2} - 1, a_{2}\right) \\
                \wedge \, \, a_{4} = ite\left(x[2] > 1, a_{3} - 1, a_{3}\right) \\
                \wedge \ldots \\
                \wedge \, \, a_{N+1} = ite\left(x[N-1] > 1, a_{N} - 1 , a_{N}\right) \\
                \wedge \neg \left(a_{N+1} \leq N\right) \\
              \end{array} \right ]  \\
              \\           
\end{equation}

The ternary operator $f \: \: ? \: \: t_1 : t_2$ shown in Figure~\ref{figure:verification-condition}(b) is converted into the conditional expression $\mathit{ite}(f, t_1, t_2)$ that takes as its first argument the Boolean formula $f$ and depending on its value selects either the second (i.e., $t_1$) or the third argument (i.e., $t_2$). In order to verify that the assertion $\left(a<=N\right)$ holds, its negation is added to $\psi$ and we check whether the entire formula is satisfiable using an off-the-self SMT solver. Formula (\ref{verification-condition-example}) can simply be represented as a Boolean logic circuit, which can further be transformed into a (equisatisfiable) CNF formula over propositional variables by Tseitin's transform~\cite{Tseitin83} in linear time and by introducing at most a linear number of fresh variables. However, checking the validity of a first-order logic formula in a given background theory is an ${\cal NP}$-complete problem~\cite{PatarinG97}).

From the practical point of view, SAT- or SMT-based BMC procedures have been successfully applied to verify a large number of hardware and software systems, including digital circuits and single- and multi-threaded programs. Those BMC techniques were able to find subtle bugs in real digital and embedded software systems, as reported in the available literature~\cite{Clarke04,MerzFS12,CordeiroF11,Ivancic05,Cordeiro12}. Nonetheless, the main criticism with respect to BMC techniques relies on completeness, since they are able to prove system correctness only if an upper bound \textit{k} is known, {\it i.e.}, a bound that unfolds all loops and recursive functions to their maximum possible depth.

Due to that limitation, BMC tools are typically susceptible to exhaustion of time or memory limits, when checking complex circuit-implementations or programs with loops, whose bounds are too large or cannot be statically determined. Indeed, such an issue has pushed researchers to overcome the depth limitation and propose extensions capable of proving global correctness.

@@

One promising approach to achieve completeness, in BMC techniques, is to prove that an invariant (assertion) is \textit{k}-inductive~\cite{EenS03,Sheera00}. The main challenge, however, of such an approach relies on computing and strengthening inductive invariants from programs. In particular, loop invariants, which are computed from programs under verification, must be inductive (and not just invariant), in order to check the corresponding verification conditions, {\it i.e.}, induction cannot determine the invariance of a non-inductive assertion~\cite{Bradley07}. 
%
For instance, consider the C-code fragment shown in Figure~\ref{figure:inductive-invariant}. If we consider the IEEE floating-point standard (IEEE 754)~\cite{4610935,Goldberg91whatevery}, then  the loop invariant $x>0$ holds initially and after each iteration; actually, $x$ tends to infinity after $128$ iterations. This invariant, however, is not inductive, given that $x>0$ before an initial iteration does not ensure that $x>0$ after each iteration; if we have initially $x=0.9$, then $x<0$ after the fourth iteration. As a consequence, even if \textit{k}-induction procedures successfully compute such assertions, which are indeed invariant, those must be inductive, so that verifiers can automatically prove program correctness. In this specific example, an inductive invariant would be $x>1$, given that if we can check that if $x>1$ before the initial iteration, then also $x>1$ after \textit{k} iterations.
%
\begin{figure}[h]
\centering
\begin{minipage}{0.45\textwidth}
\begin{lstlisting}
#include <assert.h>
int main(void) {
 float x=2;
 while (1) {
   x = ((2*x) - 1);
 }
 return 0;
}
\end{lstlisting}
\end{minipage}
\caption{Motivating example for inductive invariants.}
\label{figure:inductive-invariant}
\end{figure}

@@

\bibitem{Appel97}
Andrew~W. Appel.
\newblock {\em Modern Compiler Implementation in C: Basic Techniques}.
\newblock Cambridge University Press, New York, NY, USA, 1997.

\bibitem{Thomas01}
Thomas Ball and Sriram~K. Rajamani.
\newblock {\em SLIC: A Specification Language for Interface Checking (of C)}.
\newblock Technical Report MSR-TR-2001-21, Microsoft Research, 2001.

\bibitem{Strichman08}
Daniel Kroening and Ofer Strichman.
\newblock {\em Decision Procedures: An Algorithmic Point of View}.
\newblock Springer Publishing Company, Incorporated, 2008.

\bibitem{Tseitin83}
G.~S. Tseitin.
\newblock On the complexity of derivation in propositional calculus.
\newblock In {\em In J. Siekmann and G. Wrightson, editors, Automation of
  Reasoning 2: Classical Papers on Computational Logic 1967-1970}.

 \bibitem{PatarinG97}
Jacques Patarin and Louis Goubin.
\newblock Trapdoor one-way permutations and multivariate polynominals.
\newblock In {\em ICICS}, {\em LNCS} 1334, pages 356--368. Springer, 1997.
